Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-08-26 paper.Rnw
small changes

Numerous methods have been proposed to provide explanations for black-box model predictions \citep{gilpin:2018, guidotti:2018, ming:2017, molnar:2019}. %\hh{sounds good!}\kgc{I just took out the Mohseni reference since it is not peer reviewed. I think the other four should give enough of an idea of places to look for overviews on EML. Does that seem okay to you?}
Some are specific to one type of model (e.g. \citep{simonyan:2013} and \citep{urbanek:2008}), %\hh{good!}\kgc{I switched to a reference to Urbanek's visualizations since the Welling paper was not peer reviewed. Is this better?} %\citep{welling:2016}), 
and others are model-agnostic (e.g. \citep{fisher:2018} and \citep{strumbelj:2014}). In this paper, we focus on one specific model-agnostic method: LIME \citep{ribeiro:2016}. %introduced by Ribeiro, Singh, and Guestrin in 2016. \kgc{Do we need the addition of when the paper was introduced here since the statement is made at the beginning of Section 2, or did you intend to remove that sentence?}

@@

The concept of LIME is relatively simple: use an interpretable model to approximate a complex model in a local region. However, a practical implementation of LIME is not straightforward, and research is being done to improve the procedure \citep{laugel:2018}. The current implementations of LIME\footnote{https://github.com/marcotcr/lime}\kgc{Is there a better way to cite a Python package?}\hh{might be good to model the python package reference after the R package references, have a loog at reference limepy}  \citep{pedersen:2020} offer various tuning parameters (see Section \ref{background}) that affect the explainer model and ultimately, the explanation. Since the explainer model is an approximation of the complex model and not a direct interpretation, the explanations produced by an explainer model are subject to the quality of the approximation. Thus, in order to achieve accurate explanations, the tuning parameters selected need to be assessed.

@@

The remainder of the paper is structured as follows. Section \ref{background} provides background and claims made by \citet{ribeiro:2016} about LIME. We introduce the suggested diagnostic plots in Section \ref{diagnostics}. Then in Section \ref{application}, we demonstrate the use of the diagnostics to assess LIME explanations for a random forest  fit to a forensics bullet matching dataset. Section \ref{discussion} concludes with a discussion on extensions and limitations of the diagnostic plots and concerns about LIME in regards to the claims made by \citet{ribeiro:2016} brought about by the visualization examples in this paper that agree with \citet{alvarezmelis:2018}, \citet{laugel:2018}, and \citet{molnar:2019}. %\kgc{I moved the following two sentences here from Section 2, so that it refers to the whole paper more clearly.} \hh{I agree, that improves the flow!}

All runs of LIME in this paper are executed in a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version is functionally indistinguishable from Pedersen's implementation, but allows us to export internal values relevant for an assessment of the explainerThe diagnostic plots included in this paper are created using the R package \emph{limeaid} \citep{goode:2020}.

@@

%LIME was introduced in 2016 by \citet{ribeiro:2016}. \kgc{You added this sentence to Section 1. Do you want it removed from here or there? If it is removed here, the next sentence will need to be changed.} 
\citet{ribeiro:2016} provide an implementation of LIME in a Python package \cite{limepy}. An adaption of the Python package in R has been implemented and made available by Thomas Lin-Pedersen \citep{pedersen:2020}. The discussion of parameter choices and implementation details of LIME in this paper are based on the R package.

The general form of the LIME algorithm can be divided into three steps  \citep[see also]{laugel:2018}:

@@

The assumption of interpretability only depends on the complexity of the model used as explainer model. If the model is too complex to provide meaningful explanations (e.g. there are too many variables in the model), it is clear that the assumption of interpretability is violated. The other three assumptions are not as easy to assess, and %-- 
for those, we suggest the use of diagnostic plots.

@@

%\kgc{I moved this paragraph. Figure 3 had not been introduced yet, so I think it fits better here.}
\hh{From these scatterplots we also} see that the global relationship between response $y$ and features $x_1$ and $x_2$ is linear: the probability for label blue increases with the difference between features $x_2$ and $x_1$. Locally, the relationship between $y$ and features $x_2$ and $x_1$ varies a lot more around the line of identity. Here, the relationship is determined by the sine waves. However, the sine is a good-natured function that can be approximated well linearly in local regions.

%\kgc{I moved some paragraphs around here. It didn't read smoothly to me. I also added a sentence similar to one I had written before and you removed that I had included as an exposition sentence. Without it, I think it's confusing to jump into listing the tuning parameters we use without an overview of what is done. See what you think of it now.} 
We apply LIME using six sets of tuning parameters to all observations in the \data \ test set to observe variability across tuning parameters. A quantile bin based simulation method is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, because the complexity of the explanations increases with the number of bins. The sixth application of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications. (See Appendix~\ref{lime-details} for more details on the tuning parameters.)

For the presentation of the explanation scatterplot, we focus on the misclassified point with $(x_1, x_2, x_3)$ coordinates of (\Sexpr{round(sine_poi$x1, 2)}, \Sexpr{round(sine_poi$x2, 2)}, \Sexpr{round(sine_poi$x3, 2)}) indicated by a diamond in \autoref{fig:sine-plot-data}. Misclassified points are often of interest to explain since they may provide information about ways to improve the model. For the introduction of the other two  plots, we consider the LIME explanations for  all observations in the \data \ test data.  \kgc{XXX This is where I left off XXX}

>>>>

Numerous methods have been proposed to provide explanations for black-box model predictions \citep{gilpin:2018, guidotti:2018, ming:2017, molnar:2019}. Some are specific to one type of model (e.g. \citep{simonyan:2013} and \citep{urbanek:2008}), and others are model-agnostic (e.g. \citep{fisher:2018} and \citep{strumbelj:2014}). In this paper, we focus on one specific model-agnostic method: LIME \citep{ribeiro:2016}.

@@

The concept of LIME is relatively simple: use an interpretable model to approximate a complex model in a local region. However, a practical implementation of LIME is not straightforward, and research is being done to improve the procedure \citep{laugel:2018}. The current implementations of LIME \citep{ribeiro:2020} \kgc{Is there a better way to cite a Python package?} \hh{might be good to model the python package reference after the R package references, have a loog at reference limepy} \kgc{I kept thinking about this, and since the Python pakcgae doesn't have a manual, I ended up combining your limepy reference with the second second suggestion from https://academia.stackexchange.com/questions/14010/how-do-you-cite-a-github-repository. For now, I listed Ribeiro as the author, because he is the main contributor, but there are 40 other contributors to the package. Is it okay to just list Ribeiro?} \citep{pedersen:2020} offer various tuning parameters (see Section \ref{background}) that affect the explainer model and ultimately, the explanation. Since the explainer model is an approximation of the complex model and not a direct interpretation, the explanations produced by an explainer model are subject to the quality of the approximation. Thus, in order to achieve accurate explanations, the tuning parameters selected need to be assessed.

@@

The remainder of the paper is structured as follows. Section \ref{background} provides background and claims made by \citet{ribeiro:2016} about LIME. We introduce the suggested diagnostic plots in Section \ref{diagnostics}. Then in Section \ref{application}, we demonstrate the use of the diagnostics to assess LIME explanations for a random forest  fit to a forensics bullet matching dataset. Section \ref{discussion} concludes with a discussion on extensions and limitations of the diagnostic plots and concerns about LIME in regards to the claims made by \citet{ribeiro:2016} brought about by the visualization examples in this paper that agree with \citet{alvarezmelis:2018}, \citet{laugel:2018}, and \citet{molnar:2019}.

All runs of LIME in this paper are executed in a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (version \Sexpr{packageVersion("lime")}) by \kge{\citet{pedersen:2020}}. The forked version is functionally indistinguishable from \kge{\citeauthor{pedersen:2020}'s} implementation but allows us to export internal values relevant for an assessment of the explainer. The diagnostic plots included in this paper are created using the R package \emph{limeaid} \citep{goode:2020}.

@@

\citet{ribeiro:2016} provide an implementation of LIME in a Python package \cite{limepy}. An adaption of the Python package in R has been implemented and made available by Thomas Lin-Pedersen \citep{pedersen:2020}. The discussion of parameter choices and implementation details of LIME in this paper are based on the R package.

The general form of the LIME algorithm can be divided into three steps \citep[see also][]{laugel:2018}:

@@

The assumption of interpretability only depends on the complexity of the model used as explainer model. If the model is too complex to provide meaningful explanations (e.g. there are too many variables in the model), it is clear that the assumption of interpretability is violated. The other three assumptions are not as easy to assess, and for those, we suggest the use of diagnostic plots.

@@

From these scatterplots, we also see that the global relationship between response $y$ and features $x_1$ and $x_2$ is linear: the probability for label blue increases with the difference between features $x_2$ and $x_1$. Locally, the relationship between $y$ and features $x_2$ and $x_1$ varies a lot more around the line of identity. Here, the relationship is determined by the sine waves. However, the sine is a good-natured function that can be approximated well linearly in local regions.

We apply LIME using six sets of tuning parameters to all observations in the \data \ test set to observe variability across tuning parameters. A quantile bin based simulation method is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, because the complexity of the explanations increases with the number of bins. The sixth application of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications. (See Appendix~\ref{lime-details} for more details on the tuning parameters.)

For the presentation of the explanation scatterplot, we focus on the misclassified point with $(x_1, x_2, x_3)$ coordinates of (\Sexpr{round(sine_poi$x1, 2)}, \Sexpr{round(sine_poi$x2, 2)}, \Sexpr{round(sine_poi$x3, 2)}) indicated by a diamond in \autoref{fig:sine-plot-data}. Misclassified points are often of interest to explain since they may provide information about ways to improve the model. For the introduction of the other two  plots, we consider the LIME explanations for  all observations in the \data \ test data.
